Modeling Outbreak Frequency by Season and Pathogen: Insights and Predictions for 2025

### 2024-12-02 paper/paper.qmd
Add files via upload

Our analysis was conducted using the statistical programming language R [@citeR], leveraging a suite of packages to facilitate efficient data manipulation, modeling, and visualization. Data cleaning, transformation, and visualization were handled using `tidyverse` @tidyverse and `dplyr` @dplyr. The core analysis used Poisson regression models to explore relationships between outbreak counts and key variables, including pathogen type, seasonal variation, and outbreak settings. These models were implemented using the flexible framework offered by `rstanarm` @rstanarm for model development and extensions. Efficient data integration was achieved through arrow @arrow, and modeling workflows were supported using `modelr` @modelr within the tidyverse framework. Model summaries were generated using `modelsummary` @modelsummary, while `ggplot2` @ggplot2 was employed for creating visualizations. Enhanced table outputs were produced with `kableExtra` @kableExtra, and dynamic report generation was performed with `knitr` @citeknitr. File management, including reading parquet files, was streamlined with `here` @citehere.

@@

The cleaning process for the dataset focused on retaining key variables relevant to respiratory outbreaks in Toronto healthcare institutions. Initially, redundant or irrelevant columns were removed, and the dataset was narrowed to include only `institution_address`, `outbreak_setting`, `type_of_outbreak`, `causative_agent_1`, `date_outbreak_began`, and `active` variables. Outbreaks were filtered to focus on respiratory pathogens of interest, such as "COVID-19," "Coronavirus*," "Rhinovirus," and "Parainfluenza." The season variable was derived from the date_outbreak_began column by mapping months to their respective seasons like December-February as Winter. Invalid or missing data were addressed by removing incomplete rows or invalid entries, ensuring data consistency. These steps ensured that the dataset was streamlined for meaningful analysis while retaining critical details.

@@

# Summarize the data for plotting
plot_data <- data %>%
  group_by(causative_agent_1, season) %>% # Group by agent and season for detailed view
  summarise(total_outbreaks = sum(outbreak_count), .groups = "drop")

# Create the bar plot
ggplot(plot_data, aes(x = causative_agent_1, y = total_outbreaks, fill = season)) +
  geom_bar(stat = "identity", position = "dodge") + # Dodge for season-wise bars
  geom_text(aes(label = total_outbreaks),
    position = position_dodge(width = 0.9),
    vjust = -0.5, size = 3
  ) + # Annotate with outbreak counts
  labs(
    x = "Pathogen",
    y = "Total Outbreak Count",
    fill = "Season"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1) # Rotate x-axis labels
  )
```

@@

comparison <- data.frame(
  Metric = c("AIC", "BIC", "RMSE", "Likelihood Ratio Test (p-value)"),
  `Full Model` = c(aic_full, bic_full, rmse_full, lrt$`Pr(>Chi)`[2]),
  `Reduced Model` = c(aic_reduced, bic_reduced, rmse_reduced, "N/A") # LRT only tests nested models
)

@@

ggplot(data, aes(x = outbreak_count)) +
  geom_bar(fill = "steelblue", color = "black", alpha = 0.8) +  # Add color and transparency
  labs(
    x = "Outbreak Count",
    y = "Frequency"
  ) +  # Add labels and title
  theme_minimal(base_size = 14) +  # Use a clean theme with larger text
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold the title
    axis.title.x = element_text(face = "bold"),  # Bold x-axis label
    axis.title.y = element_text(face = "bold")   # Bold y-axis label
  ) + 
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 4)

>>>>

Our analysis was conducted using the statistical programming language R [@citeR], leveraging a suite of packages to facilitate efficient data manipulation, modeling, and visualization. Data cleaning, transformation, and visualization were handled using `tidyverse` @tidyverse and `dplyr` @dplyr. The core analysis used Poisson regression models to explore relationships between outbreak counts and key variables, including pathogen type, seasonal variation, and outbreak settings. These models were implemented using the flexible framework offered by `rstanarm` @rstanarm for model development and extensions. Efficient data integration was achieved through `arrow` @arrow, and modeling workflows were supported using `modelr` @modelr within the tidyverse framework. Model summaries were generated using `modelsummary` @modelsummary, while `ggplot2` @ggplot2 was employed for creating visualizations. Enhanced table outputs were produced with `kableExtra` @kableExtra, and dynamic report generation was performed with `knitr` @citeknitr. File management, including reading parquet files, was streamlined with `here` @citehere.

@@

The cleaning process for the dataset focused on retaining key variables relevant to respiratory outbreaks in Toronto healthcare institutions. Initially, redundant or irrelevant columns were removed, and the dataset was narrowed to include only `institution_address`, `outbreak_setting`, `type_of_outbreak`, `causative_agent_1`, `date_outbreak_began`, and `active` variables. Outbreaks were filtered to focus on respiratory pathogens of interest, such as "COVID-19", "Coronavirus*", "Rhinovirus", and "Parainfluenza". The season variable was derived from the `date_outbreak_began` column by mapping months to their respective seasons like December-February as Winter. Invalid or missing data were addressed by removing incomplete rows or invalid entries, ensuring data consistency. These steps ensured that the dataset was streamlined for meaningful analysis while retaining critical details.

@@

# Summarize the data for plotting
plot_data <- data %>%
  group_by(causative_agent_1, season) %>% 
  summarise(total_outbreaks = sum(outbreak_count), .groups = "drop")

# Create the bar plot
ggplot(plot_data, aes(x = causative_agent_1, y = total_outbreaks, fill = season)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = total_outbreaks),
    position = position_dodge(width = 0.9),
    vjust = -0.5, size = 3
  ) + # Annotate with outbreak counts
  labs(
    x = "Pathogen",
    y = "Total Outbreak Count",
    fill = "Season"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  )
```

@@

comparison <- data.frame(
  Metric = c("AIC", "BIC", "RMSE", "Likelihood Ratio Test (p-value)"),
  `Full Model` = c(aic_full, bic_full, rmse_full, lrt$`Pr(>Chi)`[2]),
  `Reduced Model` = c(aic_reduced, bic_reduced, rmse_reduced, "N/A") 
)

@@

ggplot(data, aes(x = outbreak_count)) +
  geom_bar(fill = "steelblue", color = "black", alpha = 0.8) +
  labs(
    x = "Outbreak Count",
    y = "Frequency"
  ) +  # Add labels and title
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  ) + 
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 4)
